En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Niccol\`o Veltri was supported by a research grant (13156) from VILLUM FONDEN.}

### 2019-04-30 mpc19/paper.tex
more intro

% Delay macros
\newcommand{\Delay}{\ensuremath{\mathsf{Delay}\,}}
\newcommand{\Maybe}{\mathsf{Maybe}}
\newcommand{\now}{\mathsf{now}}
\newcommand{\later}{\mathsf{later}}
\newcommand{\laterR}{\mathsf{laterR}}
\newcommand{\laterL}{\mathsf{laterL}}
\newcommand{\never}{\mathsf{never}}
\newcommand{\dn}{\downarrow}
\newcommand{\bind}{\mathsf{bind}}
\newcommand{\str}{\mathsf{str}}
\newcommand{\costr}{\mathsf{costr}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\Dapprox}{\mathbb{D}_{\approx}}
\newcommand{\copairD}[2]{[#1,#2]_{\D}}
\newcommand{\pairD}[2]{\left< #1,#2 \right>_{\D}}
\newcommand{\inlD}{\mathsf{inl}_{\D}}
\newcommand{\inrD}{\mathsf{inr}_{\D}}
\newcommand{\fstD}{\mathsf{fst}_{\D}}
\newcommand{\sndD}{\mathsf{snd}_{\D}}
\newcommand{\piso}{\mathsf{isPartialIso}}
\newcommand{\pisoalt}{\mathsf{isPartialIso}_2}
\newcommand{\iterD}{\mathsf{iter}_\D}
\newcommand{\unfolding}{\mathsf{fixpoint}_\D}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\traceD}{\mathsf{trace}_\D}
\newcommand{\traceHD}{\mathsf{traceH}_\D}
\newcommand{\daggerD}{\mathsf{dagger}_\D}
\newcommand{\Orb}[3]{\mathsf{Orb}\,#1\,#2\,#3}
\newcommand{\done}{\mathsf{done}}
\renewcommand{\next}{\mathsf{next}}
\newcommand{\reverseOrbit}{\mathsf{reverseOrb}}

@@

The coinductive delay datatype was introduced by Capretta
\cite{Capretta05} as a way of defining possibly non-terminating
functions in Martin-L\"of type theory. It has found a number of
applications, ranging from operational semantics of functional
languages \cite{Danielsson12} to formalization of domain theory in
type theory \cite{BentonKV09} and normalization by evaluation
\cite{AbelC14}. Here we employ it for giving denotational semantics to
\Pio. In particular, we show how to endow the delay datatype with a
dagger trace combinator, whose construction factors through the
specification of a uniform iteration operator
\cite{GoncharovSRJ18,GoncharovMR16}.

Throughout the paper, we reason constructively in Martin-L\"of type
theory. Classically, the delay monad is isomorphic to the maybe monad
$\Maybe\,X \equiv X + 1$,
and thus just a complication of something that can be expressed much
simpler. Constructively, however, they are very different. In particular,
it is impossible to define a well-behaved trace combinator for the
maybe monad without assuming classical principles such as the law of
excluded middle.

@@

Just as limits in $\Set$ do not lift to limits in the category $\Par$ of sets
and partial functions, they do not lift to $\Dapprox$ either. This is not an
issue with these concrete formulations of partiality, but rather with the
interaction of partiality (in the sense of restriction categories) and limits
in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is not
terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
$\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
\never : A \to \Delay 0$ as the terminal morphism. The
binary product of $A$ and $B$ in $\Dapprox$ is given by
$A + B + A \times B$. Nevertheless, it is possible to prove that 1 and
$\times$ are partial terminal object and partial binary products
respectively, in the sense of Cockett and Lack's restriction
categories \cite{CockettL02,CockettL07}. Here we refrain from making the latter
statement formal. We only show the construction of the partial pairing
operation, which we employ in the interpretation of \Pio.  Given
$f : C \to \Delay A$ and $g : C \to \Delay B$, we define:
\begin{align*}
& \pairD f g : C \to \Delay (A \times B) \\
& \pairD f g = \costr \diamond (\str \comp \pair f g)
\end{align*}
Since the delay monad is commutative, the function $\pairD f g$ is
equal to $\str \diamond (\costr \comp \pair f g)$. The operation
$\pairD - -$ is compatible with weak bisimilarity.

>>>>

% Delay macros
\newcommand{\Delay}{\ensuremath{\mathsf{Delay}\,}}
\newcommand{\Maybe}{\mathsf{Maybe}}
\newcommand{\now}{\mathsf{now}}
\newcommand{\later}{\mathsf{later}}
\newcommand{\laterR}{\mathsf{laterR}}
\newcommand{\laterL}{\mathsf{laterL}}
\newcommand{\never}{\mathsf{never}}
\newcommand{\dn}{\downarrow}
\newcommand{\bind}{\mathsf{bind}}
\newcommand{\str}{\mathsf{str}}
\newcommand{\costr}{\mathsf{costr}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\Dapprox}{\mathbb{D}_{\approx}}
\newcommand{\copairD}[2]{[#1,#2]_{\D}}
\newcommand{\pairD}[2]{\left< #1,#2 \right>_{\D}}
\newcommand{\inlD}{\mathsf{inl}_{\D}}
\newcommand{\inrD}{\mathsf{inr}_{\D}}
\newcommand{\fstD}{\mathsf{fst}_{\D}}
\newcommand{\sndD}{\mathsf{snd}_{\D}}
\newcommand{\piso}{\mathsf{isPartialIso}}
\newcommand{\pisoalt}{\mathsf{isPartialIsoCat}}
\newcommand{\iterD}{\mathsf{iter}_\D}
\newcommand{\unfolding}{\mathsf{fixpoint}_\D}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\traceD}{\mathsf{trace}_\D}
\newcommand{\traceHD}{\mathsf{traceH}_\D}
\newcommand{\daggerD}{\mathsf{dagger}_\D}
\newcommand{\Orb}[3]{\mathsf{Orb}\,#1\,#2\,#3}
\newcommand{\done}{\mathsf{done}}
\renewcommand{\next}{\mathsf{next}}
\newcommand{\reverseOrbit}{\mathsf{reverseOrb}}

@@

The full \Pio\ calculus cannot be modelled in the same framework of
\cite{CaretteS16}, since Martin-L\"of type theory is a total language
which in particolar disallow the specification of a trace
operator. Consequently, it is necessary to move to a setting
supporting the existence of partial maps, and in type theory this can
be done by using monads, by considering partiality as an effect. Our
choice fell on the coinductive delay monad, introduced by Capretta
\cite{Capretta05} as a way of representing general recursive functions
in Martin-L\"of type theory. The delay datatype has been employed in a
large number of applications, ranging from operational semantics of
functional languages \cite{Danielsson12} to formalization of domain
theory in type theory \cite{BentonKV09} and normalization by
evaluation \cite{AbelC14}. Here it is used for giving denotational
semantics to \Pio. In particular, we show how to endow the delay
datatype with a trace combinator, whose construction factors
through the specification of a uniform iteration operator
\cite{GoncharovSRJ18,GoncharovMR16}.

Throughout the paper, we reason constructively in Martin-L\"of type
theory. Classically, the delay monad is isomorphic to the maybe monad
$\Maybe\,X \equiv X + 1$,
and thus just a complication of something that can be expressed much
simpler. Constructively, however, they are very different. In particular,
it is impossible to define a well-behaved trace combinator for the
maybe monad without assuming classical principles such as the limited
principle of omniscience.

@@

Just as limits in $\Set$ do not lift to limits in the category $\Par$ of sets
and partial functions, they do not lift to $\Dapprox$ either. This is not an
issue with these concrete formulations of partiality, but rather with the
interaction of partiality (in the sense of restriction categories) and limits
in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is not
terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
$\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
\never : A \to \Delay 0$ as the terminal morphism. 
%The binary product of $A$ and $B$ in $\Dapprox$ is given by $A + B + A \times B$. 
Nevertheless, it is possible to prove that 1 and
$\times$ are partial terminal object and partial binary products
respectively, in the sense of Cockett and Lack's restriction
categories \cite{CockettL02,CockettL07}. Here we refrain from making the latter
statement formal. We only show the construction of the partial pairing
operation, which we employ in the interpretation of \Pio.  Given
$f : C \to \Delay A$ and $g : C \to \Delay B$, we define:
\begin{align*}
& \pairD f g : C \to \Delay (A \times B) \\
& \pairD f g = \costr \diamond (\str \comp \pair f g)
\end{align*}
Since the delay monad is commutative, the function $\pairD f g$ is
equal to $\str \diamond (\costr \comp \pair f g)$. The operation
$\pairD - -$ is compatible with weak bisimilarity.
